A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-03-11 analysis/paper/paper.Rmd
Accommodating multi-phased comparison

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here. A) The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black. These are plotted against the oldest radiocarbon-informed phase for each site in red. B) Using the method described by Parnell et al. (2008), the synchroneity of the dates is evaluated. When the range of the 95\\% HDR for age difference crosses zero the dates are considered synchronous.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

@@

The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use. Nonetheless, the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g. @roalkvam2022]. That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the duration and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated. Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology. However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be made explicit. This final point also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.

@@

Not least following from the fact that few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway, the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline-dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A. The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of , to allow for more deviance in the distance between site and shoreline. This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).

>>>>

```{r shoredate, echo = FALSE, fig.cap = "The result of backwards shoreline dating the sites with radiocarbon dates corresponding to the artefact inventory using the method proposed here. The shoreline dates are plotted in grey and underlined with the 95\\% HDR in black. These are plotted against the modelled radiocarbon dates for each site, which are given colour from oldest to youngest phase.", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate.png"))
```

```{r shoredatediff, echo = FALSE, fig.cap = "The agreement between the shoreline dates and radiocarbon dates given in Figure ref(fig:shoredate) are evaluated using the method described by Parnell et al. (2008). When the range of the 95\\% HDRs for age difference crosses zero the dates are considered synchronous. The colour coding follows the divison of phases given in Figure ref(fig:shoredate).", out.width = "525px", fig.align="center", regfloat=TRUE}
knitr::include_graphics(here("analysis/figures/shoredate2.png"))
```

@@

The comparison with previously reported dates is an illustrative, but unfair exercise for a few reasons. First of all the dates provided in the reports are typically stated to be a very rough estimate, and are sometimes given as a point estimate with an implied or explicit, but undefined uncertainty range. Secondly, seeing as these reports are from various dates in time, many are based on now outdated data on RSL-change. Finally, they are sometimes only meant to indicate a lower bound for when the sites could have been in use. Nonetheless, the results indicate that shoreline dating, in general, has been applied with a fairly reasonable degree of success, seeing as these dates have typically been interpreted and informed research in an approximate manner [although see e.g. @roalkvam2022]. That being said, the results do also indicate that shoreline dating has at times been applied with an exaggerated degree of precision, and while the implications of a more stable RSL-change for the duration and re-use of site locations, and consequently the precision of shoreline dating is well known, it appears to be somewhat under-appreciated. Some of the date ranges resulting from the method outlined here clearly extend well beyond major chronological divisions, even into the Iron Age, and could be severely and securely constrained with only cursory reference to typology. However, while this is obvious in some cases, the nature and uncertainty inherent to the method still means that this is arguably a required exercise that should be explicitly performed. This final point also points to the possibility of drawing on other temporal data, for example within a Bayesian framework, to further improve the precision of the dates that can be achieved with shoreline dating.

@@

Not least following from the fact that relatively few Preboreal ^14^C-dates associated with anthropogenic activity have been achieved in Norway [@åstveit2018; @damlien2018; @kleppe2018], the shoreline dating of the earliest sites is essential for understanding the pioneer settlement and the initial colonisation of the Scandinavian peninsula [e.g. @bang-andersen2012; @breivik2014; @glorstad2016]. The shoreline-dated Preboreal sites from the Brunlanes-project are among the earliest known sites in Norway [@jaksland2014a; @jaksland2012; @jaksland2012a]. These have a distinct Early Mesolithic artefact inventory and are situated in a steep area of the landscape where it would be difficult to envision use of the sites after the sea retreated any significant distance from their location. In the original publication of the sites, @jaksland2014 provides a thorough discussion of shoreline dating in general, and as used for the dating of the Brunlanes sites specifically. A comparison of his results and the ones achieved using the above-outlined approach are given in Figure \@ref(fig:brunlanes)A. The sites have been dated using what @jaksland2014 gives as the lowest elevation of finds at each site, and by employing a exponential decay ratio of , to allow for more deviance in the distance between site and shoreline. This corresponds to the decay ratio for sites older than 7000 BCE in Figure \@ref(fig:results2).
